---
title: "HW2"
author: "Haojie Liu"
date: "2023-10-17"
output: pdf_document
---

```{r}
library(ggplot2)
library(tidyverse)
library(dplyr)
library(reshape2)
```


### Problem 1: Suppose that X is a discrete random variable with probability mass function:



(a) Write R code using the inverse transform method to generate random numbers from the distribution of X.

```{r}

discrete_dist <- function(n){
  if(n<=0){
    stop("n can't be less or equal to 0")
  }
  u <- runif(n)
  x <- c()
  for(i in 1:n){
    if(u[i] < 0.1){
      x <- c(x,0)
    }else if(u[i] <= 0.3){
      x <- c(x,1)
    }else if(u[i] <= 0.45){
      x <- c(x,2)
    }else if(u[i] <= 0.75){
      x <- c(x,3)
    }else{
      x <- c(x,4)
    }
  }
  x
}


```


(b) Generate 10,000 random numbers and draw a bar chart.

```{r}

x <- discrete_dist(10000)
barplot(table(x))

```


(c) Compare the sample relative frequencies with the theoretical probability distribution.

```{r}

data.frame(x) %>% 
  group_by(x) %>% 
  count() %>% 
  summarize(relative_frequencies = n/10000) %>% 
  bind_cols(theoretical_probability = c(0.1,0.2,0.15,0.3,0.25))%>% 
  melt(id.vars="x") %>% 
  ggplot(aes(x = x,
             y = value,
             fill=variable))+geom_bar(stat="identity", position="dodge")

```


### Problem 2: Please write a function using the inverse cdf method to generate Poisson random numbers.

(a) Design an algorithm using the inverse cdf method. 

$$P(X=x) = \frac{\lambda^xe^{-\lambda}}{x!}$$
$$F(x;\lambda)=\Sigma^k_{i=0}\frac{\lambda^ie^{-\lambda}}{i!} = e^{-\lambda}\Sigma^k_{i=0}\frac{\lambda^i}{i!}$$
$$F(x;\lambda)=P(X=0)+P(X=1)+P(X=2)+...P(X=k) = e^{-\lambda} + \lambda e^{-\lambda} + \frac{\lambda^2e^{-\lambda}}{2}+...+\frac{\lambda^ke^{-\lambda}}{k!}$$
1. set $e^{-\lambda}$ be the first element of the Fx to represent the probability when k is 0.\newline
2. Generate n samples from unif(0,1)\newline
3. for each element in U, we will return the x when greater than the CDF, or add the CDF to a new px depends on the current x. 
4. return the result of the list of x.

(b) Implement your algorithm in R.

```{r}

poss_dist <- function(lambda, n){
  u <- runif(n)
  result <- c()
  for(i in 1:n){
    px <- (exp(-lambda))
    Fx <- px
    x <- 0
    while(u[i] > Fx){
      x <- x+1
      px <- (lambda^x)*(exp(-lambda))/factorial(x)
      Fx <- Fx + px
    }
    result <- c(result, x)
  }
  result
}

```


(c) Generate 10,000 random numbers with $\lambda = 4.2$ and compare your results with rpois’s.

```{r}

relative <- data.frame(relative = poss_dist(4.2, 10000))
theoretical <- data.frame(theoretical = rpois(10000,4.2))


theoretical <- theoretical %>% 
  group_by(theoretical) %>% 
  count() %>% 
  summarise(theoretical_frequancy = n/10000)
relative <- relative %>% 
  group_by(relative) %>% 
  count() %>% 
  summarise(relative_frequancy = n/10000)

relative %>% 
  left_join(theoretical, by=c("relative" = "theoretical")) %>% 
  melt(id.vars="relative") %>% 
  ggplot(aes(x = relative,
             y = value,
             fill=variable))+geom_bar(stat="identity", position="dodge")

```

Problem 3: A cumulative distribution function of X is given as following
$$F(x)=1-e^{-(x/\alpha)^\beta}, x\ge0, \alpha>0, \beta>0$$
(a) Please show that $Y = (\frac{X}{\alpha})^\beta$ follows an exponential distribution.

$$Y = (\frac{X}{\alpha})^\beta$$
Method of CDF:

$$F_Y(y) = P(Y\le y) = P((\frac{X}{\alpha})^\beta \le y)$$
$$F_Y(y) = P(X \le \alpha(y^\frac{1}{\beta})) = F_X(\alpha(y^\frac{1}{\beta}))$$
$$F_Y(y) = 1-e^{-y}$$
$$Y\sim Exp(-1)$$

(b) Write a function to generate 100,000 random numbers with $\alpha = 2$ and $\beta = 4$, and plot the histogram.

```{r}

exp_dist <- function(a, b, n){
  u <- runif(n)
  return((a*(-log(1-u))^(1/b)))
}

x <- exp_dist(2,4,100000)


hist(x, freq=FALSE, main="X")
```



### Problem 4: For the acceptance-rejection method, please prove that the returned random sample from the target density f(x).

$\frac{f(x)}{g(x)} \le M \to f(x) \le Mg(x);\forall x$$
if $u < \frac{f(x_g)}{Mg(x_g)}$ accept $x_g$ as $x_f$;
if $u \ge \frac{f(x_g)}{Mg(x_g)}$ reject $x_g$ as $x_f$

$$f(x) = \frac{P(X\; accepted\; in(x, x+\triangle x))}{\triangle x} = \frac{P(x\in (x, x+\triangle x))}{\triangle x}$$
$$P(x\in (x, x+\triangle x)) = \triangle f(x)$$

$$P(x\in (x, x+\triangle x)) = \frac{number\; of\; points\; survived \;in (x, x+\triangle x)}{accross\;the\;all\;bins,\;the\;number\; of\; points \;survived}$$

Assume that we have N as the total number of sample,

$$N * g(x)*\triangle x$$
$$P(acceptance\;for\;x\in(x, x+\triangle x)) = \frac{f(x)}{Mg(x)}$$
$$N * g(x)*\triangle x\frac{f(x)}{Mg(x)} = \frac{N}{M}\triangle xf(x)$$

Total $ of points survived:
$$\Sigma_{all\;bins}\frac{N}{M}\triangle xf(x) = \frac{N}{M}\Sigma_{all\;bins}\triangle xf(x) = \frac{N}{M}$$

Finally:

$$P(x\in (x, x+\triangle x)) = \frac{\frac{N}{M}\triangle xf(x)}{\frac{N}{M}} = \triangle xf(x)$$
$$0<\frac{f(x)}{Mg(x)} \le 1$$
$$\frac{f(x)}{g(x)} \le M$$
$$M\ge\frac{f(x)}{g(x)}\to M\ge Max(\frac{f(x)}{g(x)})$$

### Problem 5: Consider the probability mass function provided in Problem 1.

(a) Propose an envelope distribution and write R code using the acceptance-reject method to generate random numbers.

```{r}

ar_discrete_dist <- function(n){
  
  fx <- c(0.1, 0.2, 0.15, 0.3, 0.25)
  x <- 0:4
  samples <- sample(x,n,replace=TRUE)
  u <- runif(n, 0, 0.3) # envelope distribution
  result <- c()
  
  for(i in 1:n){
    if(u[i] <= fx[samples[i]+1]){
      result <- c(result, samples[i])
    }
  }
  result
}

```


(b) Generate 10,000 random numbers from the envelope distribution function. Report your empirical acceptance rate and draw a bar chart for accepted data points.

```{r}
x <- ar_discrete_dist(10000)

data.frame(table(x)/2000) %>% 
  mutate(empirical_rate = Freq) %>% 
  select(-Freq)

hist(x, probability = T)

```


(c) Compare the sample relative frequencies with the theoretical probability distribution. Discuss your choice of envelope distribution.

```{r}

length <- length(x)

data.frame(x) %>% 
  group_by(x) %>% 
  count() %>% 
  summarize(relative_frequencies = n/length) %>% 
  bind_cols(theoretical_probability = c(0.1,0.2,0.15,0.3,0.25))%>% 
  melt(id.vars="x") %>% 
  ggplot(aes(x = x,
             y = value,
             fill=variable))+geom_bar(stat="identity", position="dodge")

```

We chose a uniform envelope distribution for simplicity, given that we were working with a discrete set of values. However it will be better to choose the distribution that is looks more likely to the real distribution.


### Problem 6: Write a function to generate random variables from the $Beta(\alpha, \beta)$ distribution using the acceptance-rejection method. You may use Unif(0, 1) as the envelope distribution. You may set $\alpha = 2$ and $\beta = 3$.

(a) Calculate M before coding.

$$f(x) = \frac{x^{\alpha -1}(1-x)^{\beta-1}}{B(\alpha,\beta)} = \frac{x(1-x)^2}{1/12}$$
$$g(x) \sim unif(0,1) = 1$$
$$M = max(\frac{f(x)}{1})= max(12x(1-x)^2)$$
The function become maximum at $x=1/3$, therefore $M\simeq1.78$

(b) Generate a random sample of size 100,000 from Uniform(0, 1), and plot the histogram for accepted data points.

```{r}

beta_dist <- function(a, b, n){
  

  x <- runif(n)
  u <- runif(n)
  result <- c()
  fx <- function(x){
    M <- 1.78
    return((x^(a-1)*(1-x)^(b-1))/(M*beta(a,b)))
  }
  
  for (i in 1:n) {
    if(u[i] <= fx(x[i])){
      result <- c(result, x[i])
    }
  }
  
  result
  
}

hist(beta_dist(2,3,100000), freq=FALSE, 
     main = "Beta Distribution with (alpha = 2, beta = 3)",
     xlab="x")


```

(c) Compute the empirical acceptance rate, and compare it with 1/M.

```{r}

x <- beta_dist(2,3,100000)

data.frame(empirical_rate = length(x)/100000, theoretical_rate = 1/1.78)

```


### Problem 7: The standard Laplace density is $$f(x)=\frac{1}{2}e^{-|x|},x\in \mathbb{R}$$

(a) Design an algorithm to generate 10,000 random variables using the inverse CDF method and implement it in R.

$$F(x) = \int^x_0\frac{1}{2}e^{-t}dt = \frac{1}{2}(-e^{-x}+1); \forall x\ge0$$
$$F^{-1}(U) = \ln(2U)$$

$$F(x) = \int^x_{-\infty}\frac{1}{2}e^{t}dt = \frac{1}{2}(-e^{x}+1); \forall x<0$$
$$F^{-1}(U) = -\ln(2-2U)$$

```{r}

inv_laplace_dist <- function(n){
  u <- runif(n)
  neg <- log(2*u[u<0.5])
  pos <- -log(2-2*u[u>=0.5])
  
  return(c(neg,pos))
}

hist(inv_laplace_dist(10000))

```


(b) Design an algorithm to generate 10,000 random variables using the rejection method and implement it in R. You may use Normal(0, 3) as the envelope distribution.

$$M = max(\frac{f(x)}{g(x)}) = max(\frac{\frac{1}{2}e^{-|x|}}{N(0,3)})$$

1. Define the function `ap_laplace_dist(n)`:
    1.1 Input: `n` (integer) - the number of samples to generate.

2. Define the function `fx(x)`:
    2.1 If `x >= 0`, return `0.5 * (-exp(-x) + 1)`.
    2.2 If `x < 0`, return `0.5 * (-exp(x) + 1)`.

3. Generate an array `x` of `n` samples from a normal distribution with mean `0` and standard deviation `3`.

4. Calculate `M` as the maximum value of the following: 
    4.1 For each `i` in the range from `-5` to `5` with a step of `10000`, calculate `fx(i) / x[i]`.

5. Generate an array `u` of `n` samples from a uniform distribution between `0` and `1`.

6. Initialize an empty list `result`.

7. For `i` from `1` to `n`:
    7.1 If `u[i] <= fx(x[i])`, append `x[i]` to `result`.

8. Return `result`.


```{r}

ap_laplace_dist <- function(n){
  
  fx <- function(x){
    if(x>=0){
      return(0.5*(-exp(-x)+1))
    }else{
      return(0.5*(-exp(x)+1))
    }
  }
  
  x <- rnorm(n, 0, 3)
  M <- max(sapply(seq(-5, 5, length.out = 10000), fx)/x)
  u <- runif(n)
  result <- c()
  
  for (i in 1:n) {
    if(u[i] <= fx(x[i])){
      result <- c(result, x[i])
    }
  }
  
  result
  
}

hist(ap_laplace_dist(10000))


```


(c) Compare your results of (a) and (b). Discuss their advantages and disadvantages.

The advantages of method in (a) is that it will estimate the sample distribution more precisely, but with a longer calculation process. The advantages of method in (b) is that it will take a shorter time to estimate the sample distribution via we don't really know the real CDF, however, the correctness of estimation will base on the envelope distribution that we choose. 








