---
title: "HW3"
author: "Haojie Liu"
date: "2023-11-03"
output: pdf_document
---

```{r message=FALSE}
library(tidyverse)
library(ggplot2)
library(reshape2)
```


### Problem 1: Given the 4-dimensional multivariate normal distribution with mean vector 
$$\mu^T = \begin{pmatrix}2 &1.5&3&1 \end{pmatrix}$$
and covariance matrix
$$\Sigma = \begin{pmatrix}2.8&0&0.2&2\\0&1.7&2&0\\0.2&2&3.6&-1.2\\2&0&-1.2&3\end{pmatrix}$$
(a) Generate 1000 random observations from this multivariate normal distribution using the Choleski factorization method.

```{r}
mu <- c(2,1.5,3,1)
covmat <- matrix(c(2.8,0,0.2,2,0,1.7,2,0,0.2,2,3.6,-1.2,2,0,-1.2,3), nrow = 4)
x <- matrix(rnorm(4000),ncol = 4)
sample <- matrix(numeric(4000), nrow = 1000)
for(i in 1:1000){
  sample[i,] <- mu+(x%*%chol(covmat))[i,]
}
```


(b) Draw an array of scatter plots for each pair of variables and examine if they agree with the parameters. (You may use pairs in R)

```{r}

pairs(sample)

```

Base on the pairs plot, I can tell that all means are around the $\mu$ vector that was given, and similar $\Sigma$ that was given.


### Problem 2: Write R code to standardize an d-dimensional multivariate normal sample X with the known $\Sigma$ and sample size n, where d and n can be arbitrary integer numbers.
(a) Derive an algorithm for standardizing a multivariate normal sample. 

To make the multivariate normal sample become standardized, we will make each sample subtract by their mean and divide by the standard deviation:
$$x_{standarized} = \frac{x - \mu}{\sigma}$$
Or in this example we have d-dimensional multivariate normal X with size n

$$X_{standarized} = (X-\mu)\Sigma^{-\frac{1}{2}}$$

(b) Implement your algorithm in R.

```{r}
standarize <- function(X, mu, covmat){
  
  df <- matrix(numeric(length(X)), nrow = nrow(X))
  for(i in 1:nrow(X)){
    df[i,] <- X[i,]-mu
  }
  
  eigen_value <- eigen(covmat)
  covmat_inverse <- eigen_value$vectors %*% diag(1/sqrt(eigen_value$values)) %*% t(eigen_value$vectors)
  
  return(df %*% covmat_inverse)
}
```


(c) Use the generated data from Problem 1 to verify your algorithm.

```{r}
df <- standarize(sample, mu,covmat)
pairs(df)
```
Since all of the scatter plots looks randomly spread around mean equals 1, we can verify that the algorithm works at standarize the data from problem 1.


### Problem 3: Given X is a continuous random variable from the density f(x). Let $\theta = \int g(x)f(x)dx = E[g(x)]$. Suppose we draw iid samples $X_1,...,X_m$ from $f(x)$. Let $\hat \theta = \frac{1}{m}\Sigma_{i=1}^mg(X_i)$.

(a) Prove $E[\hat \theta] =\theta$

$$E[\hat \theta] = E[\frac{1}{m}\Sigma_{i=1}^mg(X_i)] = \frac{1}{m}E[\Sigma_{i=1}^mg(X_i)]$$
$$E[\hat\theta] = \frac{1}{m}\Sigma_{i=1}^m\theta = \frac{m\theta}{m} = \theta$$

(b) Prove $Var[\hat \theta] = Var[g(X)]/m$, and specify how to estimate $Var[g(x)]$

$$Var[\hat \theta] = Var[\frac{1}{m}\Sigma_{i=1}^mg(X_i)] = \frac{1}{m^2}Var[\Sigma_{i=1}^mg(X_i)]$$
$$Var[\hat \theta] = \frac{1}{m^2}(Var[g(X_1)]+...Var[g(X_m)]) = \frac{1}{m^2}mVar[g(X)] = \frac{Var[g(X)]}{m}$$

(c) Specify how to construct 99% confidence interval of $\theta$ using central limit theorem.

$$CI = \hat\theta \pm(z_{\alpha/2}\times SE)$$
$$CI = \frac{1}{m}\Sigma_{i=1}^mg(X_i)\pm(2.576\times\sqrt{\frac{Var[g(X)]}{m}})$$

(d) Suppose $f(x)$ is the exponential density with the rate, 1/3. Write a function (`mc2()`) to calculate a Monte Carlo estimate of $E[\sqrt{X}]$

```{r}

mc2 <- function(m){
  
  lambda <- 1/3
  u <- runif(m)
  exp <- -log(1-u)/lambda
  return(c(mean(sqrt(exp)), sd(sqrt(exp))))
  
}

```


(e) Construct the 95% confidence interval of $E[\sqrt{X}]$. Repeat your function 1000 times, how often the confidence interval capture the true value of $E[\sqrt{X}]$.


```{r}
means <- replicate(1000,mc2(1000))
true_mean <- integrate(function(x) sqrt(x) * 1/3 * exp(-1/3 * x), lower = 0, upper = Inf)$value
CI <- list(means[1,] + qnorm(0.975)*means[2,], means[1,] - qnorm(0.975)*means[2,])
result <- logical(0)
for (i in 1:1000) {
  if(true_mean>CI[[1]][i]){
    result <- c(result,FALSE)
  }else if(true_mean<CI[[2]][i]){
    result <- c(result,FALSE)
  }else{
    result <- c(result,TRUE)
  }
}

mean(result)
```

Since all we have all TURE inside the result vector, we can tell that all of the estimated confidence interval captures the theoretical true mean.

### Problem 4: Find the air-conditioning data set aircondit from the boot package. The data includes the 12 time intervals in hours between successive failures of the air-conditioning equipment. Assume that the time intervals between failures follow an exponential distribution with the hazard rate $\lambda$. Please use bootstrap to estimate the bias and standard error of $\hat\lambda_{MLE}$.

```{r}
library(boot)

# By MLE, we can estimate our lambda as the following:
MLE_lambda <- function(df, index){
  resampled_data <- df[index, ]
  return(1/mean(resampled_data))
}


boot(aircondit, MLE_lambda, R = 10000)
```


### Problem 5: Suppose X is a random variable from Beta($\alpha$ = 3, $\beta$ = 2).
(a) Write R code to compute the Monte Carlo estimator of the CDF. 

$$f(x) = \frac{x^{\alpha -1}(1-x)^{\beta-1}}{B(\alpha,\beta)} = \frac{x^2(1-x)}{1/12}$$

$$F(x) = \int^x_012t^2(1-t) dt = 4x^3-3x^4$$
$$u = 4x^3-3x^4$$


```{r}

simulated_values <- rbeta(5000, 3, 2)

estimate_cdf <- function(x) {
  mean(simulated_values <= x)
}

mc_estimates <- sapply(seq(0.1, 0.9, by = 0.1), estimate_cdf)
hist(mc_estimates)

```


(b) Write R code using the “hit-or-miss” approach to estimate the CDF.


```{r}

hit_or_miss <- function(x, alpha, beta, n) {
  hits <- 0
  for (i in 1:n) {
    u <- runif(1)
    y <- dbeta(u, alpha, beta)
    v <- runif(1, 0, max(dbeta(seq(0, 1, length.out = 1000), alpha, beta)))
    if (v <= y) {
      hits <- hits + (u <= x)
    }
  }
  return(hits / n)
}

hit_or_miss_estimates <- sapply(seq(0.1, 0.9, by = 0.1), hit_or_miss, alpha = 3, beta = 2, n = 1000)
hist(hit_or_miss_estimates)

```


(c) Compare your estimates with the outputs of the pbeta function in R for x = 0.1, 0.2, ..., 0.9.


```{r}

exact_cdf <- pbeta(seq(0.1, 0.9, by = 0.1), 3, 2)

comparison <- data.frame(
  x = seq(0.1, 0.9, by = 0.1),
  Monte_Carlo = mc_estimates,
  Hit_or_Miss = hit_or_miss_estimates,
  Exact = exact_cdf
)

comparison %>% 
  melt(id.vars = "x")%>% 
  ggplot(aes(x = x,
             y = value,
             fill = variable))+geom_bar(stat="identity", position="dodge")

```


















