---
title: "Note 4"
author: "Haojie Liu"
date: "2023-10-10"
output: pdf_document
---

From last time
$$F_X(x), u\sim unif(0,1)$$

$$F^{-1}(u)= inf\{x: F_X(x) = u\}$$

$$F^{-1}(u) = inf\{x_i:F_X(x_{i-1}) < u\le F_X(x_i)\}$$


Geometric distribution $x = 0,1,...$

$$F^{-1}(u) = inf\{x:F_X(x) < u \le F_X(x+1)\}$$

$$\frac{log(u)}{log(q)}\le x+1$$
ceiling function

$$\lceil\frac{log(u)}{log(q)}\rceil - 1=x$$

Log distribution function

$$P(x) = \frac{\theta^x}{-log_{10}(1-\theta)x}; x =1,2,...;0<\theta<1$$

number of items purchased by a customer in a unit of time.


$$F(x) = P(1)+P(2)+...+P(x)$$

$$F_X = \begin{pmatrix}F(1)\\F(2)\\:\\F(x)\\F(x+1)\\:\end{pmatrix} = \begin{pmatrix}I(u\ge F(1))\\I(u\ge F(2))\\:\\I(u\ge F(x))\\I(u\ge F(x))\\:\end{pmatrix} = I_F$$

Generate $u\sim Unif(0,1)$ cheich if $u<F_X(x)$ by order $x = 1,2,...$ until $u\ge F_X(x) \to$ return x

**Ex:**\newline

if $u<F(1)$

$$I_F = \begin{pmatrix}F\\F\\F\\:\\F\\F\end{pmatrix}$$ 
$$sum(I_F)= 0 + 1$$

if $u<F(4)$

$$I_F = \begin{pmatrix}T\\T\\T\\:\\F\\F\end{pmatrix}$$ 
$$sum(I_F)= 3 + 1$$

$$F_X(x) = \Sigma^x_{k=1}P(x) = F(x-1) +P(x)$$
$$u\sim unif(0,1)$$

$$while(u>F_X(x))\{\\  
  x = x+1\\
  Fx = Fx[x-1] + px
  \}\\return x$$
```{r}
logdist <- function(u, Fx){
  
  while(u>Fx){
    x <- x+1
    newFx <- Fx + px
  }
  return(x)
}
```

Recursive equation

$$P(x) = rec(P(x-1))$$

$$P(x=1)= \frac{\theta}{-log(1-\theta)}$$
$$P(x=2)= \frac{\theta^2}{-log_{10}(1-\theta)2} = \frac{\theta}{-log_{10}(1-\theta)}\frac{\theta}{2} = P(x=1)\frac{\theta}{2}$$
$$P(x=3)= \frac{\theta^3}{-log_{10}(1-\theta)3} = \frac{\theta}{-log_{10}(1-\theta)}\frac{\theta}{2}\frac{2\theta}{3} = P(x=2)\frac{2\theta}{3}$$
$$P(x) = P(x-1)\frac{(x-1)\theta}{x}$$

$$rlog(\theta)\{\\ x = 1\\ p(x) = \frac{\theta}{-log_{10}(1-\theta)}=F(x)\\u\sim unif(0,1)\\while(u>F_X(x))\{\\x = x+1\\P(x)=P(x-1)\frac{(x-1)\theta}{x}\\F(x)=F(x-1)+P(x)\\\}\\return(x)\}$$

```{r}
logdist <- function(theta,n){
  x <- 1
  px <- theta/-log((1-theta),10)
  Fx <- px
  u <- runif(n)
  while(u>Fx){
    x <- x+1
    newpx <- px*((x-1)*theta)/x
    newFx <- Fx + px
  }
  return(x)
}

```


Some of the function that I can't be able to use the inverse CDF method, since it does not have a close form.

$$f(x) = \frac{1}{(2\pi)^2}e^{-\frac{x^2}{2}}$$
This is a example


## Acceptance-Rejection Method (indirect method)

We will find a g(x) to generate samples

$$g(x)\to x_1^{g}...x_n^{g}$$

Then will create a mechanism to determine if $x_i^g$ is accepted $x\sim f(x)$

Constraints:

$$1. f(x)> 0, g(x)>0$$
$$2. \frac{f(x)}{g(x)} \le M_{const} \to f(x)\le g(x)M_{const}; \forall x$$

$$x_g \sim g(x); x_f \sim f(x)$$

if $\frac{f(x_g)}{Mg(x_g)}$ is small it will be less likely from $f(x)$
if $\frac{f(x_g)}{Mg(x_g)}$ is large it will be more likely from $f(x)$






